Rheology of Cubic Blue Phases

### 2012-08-28 papers/bprheo/bprheo.tex
Test commit
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1446 40876441-3496-0410-a613-a2fca35b2ae2
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1763 40876441-3496-0410-a613-a2fca35b2ae2

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,pre,twocolumn,groupedaddress]{revtex4-1}
%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}

@@

@@

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.
\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
\affiliation{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}

%\email[]{}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}

@@

\begin{abstract}
\end{abstract}

% insert suggested PACS numbers in braces on next line
\pacs{}
% insert suggested keywords - APS authors don't need to do this
%\keywords{}

@@

In the following we report results on the bulk flow behaviour of BPI and BPII for thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
For these parameters both blue phases are equilibrium phases in the quiescent state.
We have also performed simulations at higher temperatures and chiralities and for metastable configurations, which led to very similar results.
Quantitative differences occured only in that sense that the disclination network dissolved more easily for a particular flow rate if the system was closer to the cholesteric-isotropic phase boundary and where the order parameter takes lower values and free energy density is higher.\\
The tensor order parameter was initialized with the high-chirality solution of Eq. \ref{eqn1}, which has the correct topology of the actual blue phase.
The initial configurations were equilibrated for $5000$ LB time steps, where we allowed for optimization of the unit cell size \cite{Alexander:2006}, a feature that is often referred to as 'redshift'.
After the equilibration sequence the unit cell size was fixed and the shear flow was switched on.
For BPII we used a pitch length of 32 lattice sites, but in case of BPI it was necessary to increase the resolution to 64 sites per pitch length to get accurate results of the flow induced transformations in the disclination network.
In order to impose a flow profile we used Lees-Edwards boundary conditions \cite{Wagner:2002}, whereas the top (bottom) half of planes moved in positive (negative) y-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross the Lees-Edwards planes \cite{Henrich:2011b}.
The imposed flow rates ranged over two orders of magnitude from about $1.25\cdot 10^{-6}$ to $6\cdot10^{-4}$ in LB units.\\
In the visualizations that follow the orientation of the principal coordinate axes in x-, y- and z-direction always coincides with the direction of the velocity gradient direction, the flow and the vorticity, respectively.
Disclination lines are plotted as isosurface of the scalar order parameter $q$

@@

We start the discussion with BPII, because its flow behaviour is simpler and probably more easily anticipated.
Fig \ref{fig1} shows the disclination network undergoing a flow-induced transformation.
The periodically recurring structure remains very regular and apart from the quasi-affine transformation relatively close to its quiescent appearance with individual disclination lines breaking up and reconnecting further downstream.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\cdot 10^{-4}$. The horizontal y-direction is the flow direction whereas the velocity gradient is oriented along the vertical x-direction.}
\label{fig1}
\end{figure*}

Interestingly, while being steadily sheared in y-direction, the entire disclination network is slowly displaced in z-direction, the direction of vorticity.
This feature is most clearly observed when the structure is viewed along the flow direction.
Fig. \ref{fig2} depicts a sequence of close-ups of the disclination network as it travels in negative z-direction, i.e. from the foreground towards the background of the picture. 
The velocity $v_z$, represented by the colour code, exhibits a recurring pattern of positive and negative values.
Peak velocities occur close to disclination lines where the molecular order is suppressed or in regions where two disclination lines approach each other and the network is about to break up or reform.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-149k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-160k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-167k_run902.png}\\
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-173k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-182k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-192k_run902.png}\\
\caption{A section of $2\times2\times1$ unit cells of BPII seen along the direction of vorticity. The colour code gives the magnitude and direction of the velocity component $v_z$ perpendicular to the plane.}
\label{fig2}
\end{figure*}

Plotting only x- and z-component of the local velocity allows to visualize the spatial dependence of the secondary flow in the blue phase with respect to the disclination network.
Fig. \ref{fig3} shows characteristic regions of upward and downward 'flow', which are normally superimposed by the primary shear flow in y-direction.
Remarkably, the secondary flow is directly connected to the sense of the underlying cholesteric helix as changing from left- to right-handed helix also inverts the sense of motion of the network.
%%%%%%%%%%%%%%%%%%%%

The lefthand picture depicts the situation the movement of the disclination network is downwards, while helix inversion leads to an upward motion in the righthand picture.
More precisely the orientation of the velocities with respect to the bands is exactly inverted in both cases.

A closer look at the shape of the disclination lines reveals that also the network seems to be inverted. 
The total magnitude of the secondary flow is in the range of a few percent of the primary flow and the images in Fig. \ref{fig3} give actually the situation close to the peak point.\\

A secondary flow pattern with distinctive regions of positive and negative velocity components $v_z$ emerges and reoccurs periodically.

\begin{figure*}[h]
\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run902.png}
\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run903.png}
\caption{Secondary flow in BPII visualized by means of the velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity of the cholesteric helix. The images depict an individual snapshot of the periodically recurring sequence in the steady state.}
\label{fig3}
\end{figure*}

It is interesting to study the flow behaviour from a more quantitative point of view by looking at the shear stress dependence on flow rate and time.

The effective viscosity $\eta_{eff}$ over time is shown in Fig. \ref{fig4}. 
For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig1}.
However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
The inset shows the deviatoric part of the chemical stress tensor as a function of shear rate.
The dependence is that of a power law $\sigma_{eff, xy}=a \dot{\gamma}^b$ with $a=0.34(6116), b=0.95(2654)$ for low and intermediate shear rates. 
The error bars indicate the maximum and minimum stress during one cycle.
The idependence is almost linear and gives further evidence of the remarkably small degree of shear-thinning and also reflected in the very regular and undisturbed affine transformation of the disclination network.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{stress_bp2.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPII in the steady state. }
\label{fig4}
\end{figure}

Above a critical shear rate, which is $\dot{\gamma}=3.1-6.2\cdot 10^{-4}$ in our case, the rotational diffusion coefficient $\Gamma$ in Eq. \ref{eqn2} is too small and the disclination network adopts a completely different conformation in the steady state.
Fig. \ref{fig5} shows the director field and the position of disclination plane.
The degree of twist is the same as that of the underlying cholesteric, but the orientation of the director on either side of the planes differs by almost 90 degrees, giving rise to four distinct layers, which are separated by disclination planes.
Even in this flow-induced state the disclinations move and retain the above described sense of motion in the direction of vorticity.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{dir-op-y_225k_run949.png}
\caption{BPII at high shear rate: director field and the disclination planes seen along the flow direction. The BPII breaks up and flows in helical layers separated from each other by a disclination plane}
\label{fig5}
\end{figure}

@@

The flow behaviour of BPI deviates significantly from that of BPII in a number of points.
First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig1}, but this time seen along the direction of flow.
A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
The initial and final state in this sequence correspond to half a cycle. 
This can be easily verified by comparing the position of the disclination network at the left or right boundary.
BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI seen along the flow direction in steady shear flow at $\dot{\gamma}=1.56\cdot 10^{-4}$, showing the downward movement of the disclination network in the direction of vorticity.}
\caption{\label{fig6}}
\end{figure*}

The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.

\begin{figure*}[h]
\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run914.png}
\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run922.png}
\caption{Secondary flow in BPI: velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity.}
\label{fig7}
\end{figure*}

Despite qualitative similarities between the BPI and BPII, the quantitative analysis reveals striking differences.
In Fig. \ref{fig8} we show the effective viscosity $\eta_{eff}$ over time and flow curves $\sigma_{chem}(\dot{\gamma})$. 
BPI exhibits pronounced shear-thinning contrary to BPII, whereas the shape of the curves in no longer symmetric.
In fact on approaching the critical shear rate that led in the case of BPII to the disruption of the network, the curves become increasingly complex but remain fully periodic.
The effective viscosity drops by about an order of magnitude between the lowest and the highest shear rates.
Note that the scale has been adapted to allow for the different magnitudes.
The inset can be directly compared with that of Fig. \ref{fig4}.
Interestingly the difference between minima and maxima of the chemical shear stress is much larger than for BPII and for the lowest shear rate we applied the absolute minimum takes even negative values, which are not shown in Fig. \ref{fig4}.
Our understanding of this phenomenon is an interplay of the elastic forces and the individual order paramter configuration of the liquid crystal.
If the system is close to a local minimum of the free energy but moving away from it the increasing strain pumps elastic energy into the system. 
This results in positive chemical stresses as the system tries to withstand the deformation.
However, if the system has sufficiently transformed and is about to overcome a local maximum, it will be pulled towards the next local minumum, which can lead temporarily to negative chemical stresses. 
The inset shows that the chemical stress decreases monotonically with decreasing flow rate, but does not follow a simple power law dependence.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{stress_bp1.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPI in the steady state.}
\label{fig8}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{fig9}
\end{figure}

Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{dir-op-z_395k_run916.png}
\includegraphics[width=0.5\textwidth]{dir-op-x_395k_run916.png}
\caption{BPI at high shear rate: director field and the disclination planes seen along the direction of voritcity and the gradient direction. BPI forms rolls with $\pi/2$ disclinations that are advected with the general flow.}
\label{fig10}
\end{figure}

>>>>

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,pre,reprint,superscriptaddress]{revtex4}
%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}

@@

\newcommand{\e}[1]{\times10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

@@

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.

\author{O. Henrich}
\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}

\author{K. Stratford}
\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{P.V. Coveney}
\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}

\author{M.E. Cates}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{D. Marenduzzo}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}

@@

\begin{abstract}
We study the behaviour of cubic blue phases under shear flow
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
\end{abstract}

% insert suggested PACS numbers in braces on next line
\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
% insert suggested keywords - APS authors don't need to do this
%\keywords{}

@@

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
We opted for states sufficiently far away from the chiral-nematic-isotropic 
where both quiescent blue phases are at equilibrium, namely
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
Simulations at metastable states at higher and lower temperatures and different chiralities showed 
no differences in their generic flow behaviour.
However, states closer to the phase boundary featuring higher free energy densities and 
smaller average order parameters show a tendency to break up more easily under external driving forces.

As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
the complex conformational changes and kinematic details of the blue phase networks.

Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
along the y- and z-axis.
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa.
More details about the conversion from LBU to SI units can be found in
\cite{Henrich:2011a,Henrich:2010b}.

@@

We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
intersect and form a characteristic network of nodes.

\subsubsection{Low and intermediate shear rates}

Fig \ref{bp2-med} shows the disclination network 
undergoing an affine transformation under shear. The disclination lines break up and 
reconnect further downstream forming a periodically recurring pattern. 
The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
Apart from the affine transformation the general appearance is very close to that 
of the quiescent equilibrium phase.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions impose flow in such a way that network moves 
to the right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure*}

Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the z-direction.
The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.

A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
the picture on the right shows the same data versus accummulated total strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is due to the periodic break-up and reconnection of the network under shear flow.

The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
The error bars indicate the maximum and minimum stresses that occur during one cycle.
The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
a very good approximation, exibiting a remarkably small degree of shear-thinning.
This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure*}

Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
and the sense of motion also changes.
Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocity in z-direction.
Hence the movement of the network in vorticity direction is predominantly permeative.
Note that the imbalance is exaclty retained if the helicity is inverted.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

\subsubsection{High shear rates}

If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
shear rates \cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
Due to a lack of computational resources we have not investigated these aspects any further.

\begin{figure}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The configuration is a helical wave that travels along the direction of vorticity. It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure}
\clearpage

@@

In BPI the disclination lines are arranged in a body-centered-cubic array.
Contrary to BPII, they are well separated and do not intersect. 
This difference in topology leads to some stiking differences in the flow behaviour of both blue phases 
as we will see in the following.

We can identify three different regimes in the range of applied shear rates:
At high shear rates a breakup of the network structure occurs, similar to that seen in BPII. 
At intermediate shear rates BPI features seemingly chaotic but periodically recurring conformations.
At low shear rates the network undergoes first an affine transformation, but becomes unstable
at larger strains and adopts an amorphous configuration with well separated disclination lines.

\subsubsection{Intermediate shear rates}

Fig. \ref{bp1-med} shows a typical sequence of shear-induced conformations.
The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.

There is again an offset in the position of the network at the begin and the end of each cycle. 
But neither the magnitude nor the direction of the offset seem to obey a clear rule.
They are seemingly chaotic and different for every shear rate.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. The network is displaced during every cycle, but the magnitude and direction of the offset are seemingly chaotic.}
\label{bp1-med}
\end{figure*}

The apparent viscosity of BPI in simple shear flow, depicted in Fig. \ref{bp1-rheo}, is larger for The degree of shear thinning, also visible in the inset in terms of average shear stress over
flow rate

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp1-rheo}
\end{figure}

Runs at smaller system sizes suggest that the individual conformation 
can depend sensitively on the system size. Fig. \ref{bp1-2uc4uc}

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{}
\label{bp1-2uc4uc}
\end{figure}

\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N)
\end{equation}

@@

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{bp1-spec}
\end{figure}

%%%%%%%%%%%%%%%%%%%
The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero. The colour code gives the magnitude and sign of the z-component. The vertical and horizontal direction are the gradient and vorticity direction, respectivelty. Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
\label{bp1-velo}
\end{figure*}

\subsubsection{Low shear rates}

At low shear rates the conformational changes of BPI in shear flow are very similar to 
that of BPII descibed above. Fig. \ref{bp1-low} shows how the network undergoes an affine 
transformation during one cycle. However, contrary to BPII the disclination lines of BPI 
do not resemble those of the quiescent state, although they have the same topological
structure. At one particular point (top-left and top-centre) 
the disclination network are intertwined helices with the helical axis aligned
along the gradient direction. This shear-induced conformation forms shortly after the
flow has been switched on at strains $\gamma\simeq1$.
During one cycle there is no perceivable movement of the network in vorticity direction.

\begin{figure}[h]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Disclination network of BPI at low shear rate: Depicted is a sequence of affine, shear-induced transformations at shear rate $\dot{\gamma}=4.88\e{-6}$ and time step $t=4.0, 4.0, 6.0, 7.25, 7.50, 7.75\e{5}$ (top-left to bottom-right). The horizontal and the vertical direction are the flow and the gradient direction in all but the top-centre picure. There, the network is depicted in gradient direction with the direction of vorticity oriented along the vertical axis.}
\label{bp1-low}
\end{figure}

\subsubsection{High shear rates}

\begin{figure}[h]
\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis. The colour code gives the magnitude of the in-plane component of the director field. The data is translationally invariant along the vorticity direction.}
\label{bp1-high}
\end{figure}

\clearpage
